Variation in distance between consecutive animal locations

### 2021-04-15 distributions_paper/rudolph_distributions.Rmd
Progress with methods

  We focus here on developing a process to analyze animal movement sequences based on movement lengths, or step lengths, and turning angle probability distributions. We assume the movement is a correlated random walk and derive expected square displacements associated to a number of consecutive moves under various probability distribution models. Unlike other studies which focus on deciphering the underlying behaviors from animal movement data, our focus was to uncover potential differences between individuals and how these differences can be obscured when modeling population-level animal movement. We know these simple correlated random walks, or simple movement models might not be enough to describe net displacements under more complex processes, for which developments in higher order markov processes show to be promising. However, our goal in this study was to use a simpler model and focus on potential differences that arise when comparing data from multiple individuals under frameworks that completely pool data in contrast to those that don't.

@@

## Animal movement model

As in other papers. Note that we adopt the convention of using upper case letters for random variables and lower case letters for possible numerical values of these random variables.

## Description of the models considered

## Specific case with telemetry data

>>>>

Animal telemetry data are collected by determining an animal's location at various points in time. Many of the movement models used to analyze animal telemetry data rely on regular time intervals between locations [@jonsen2003meta; @jonsen2005robust]. However, based on field methods, or environmental variables such as heavy cloud cover or dense forest, can make data collection at regular intervals difficult, and thus location data becomes more opportunistic and with irregular time intervals. Previous methods to analyze data under these irregular conditions have relied on interpolating, subsampling, or aggregating data during pre-processing in order to make it fit model assumptions[@johnson2008continuous]. In this study, we propose the use of a model capable of handling irregular time intervals by considering movement as a stochastic process on a continuous time scale, and thus focus on the animal's instantaneous velocity rather than on its absolute displacements between locations. Using this continuous-time framework, we are able to estimate movement paths by integrating velocity functions. We also explore the effects of individual variation in speeds as a comparison between velocity probability density functions across different levels of data aggregation between individuals. Unlike other studies which focus on deciphering the underlying behaviors from animal movement data, our focus was to uncover potential differences between individuals and how these differences can be obscured when modeling population-level animal movement. We know these simple correlated random walks, or simple movement models might not be enough to describe net displacements under more complex processes, for which developments in state-space modeling frameworks show to be promising. However, our goal in this study was to use a simpler model that would adjust to irregular and sparse telemetry data, and focus on potential differences that arise when comparing data from multiple individuals under frameworks that completely pool data in contrast to those that don't. We illustrate the use of this model and the individual variation comparison by analyzing a telemetry data set of twelve many-banded aracari individuals, *Pteroglossus pluricinctus*.

@@

## Velocity-based continuous-time model

We developed an animal movement model with telemetry data in mind, thus we considered a movement trajectory as a series of consecutive recordings of an animal's location. To first set up the framework for a velocity-based model, we consider movement as the change in locations across two coordinates. We let $\mu(t)=[\mu_x(t), \mu_y(t)]'$ be the location of an individual at time $t$ in the coordinates $x$ and $y$. We establish $\boldsymbol{d}_{\Delta}(t) = \boldsymbol{\mu}(t+\Delta) - \boldsymbol{\mu}(t)$ as the difference in locations across both coordinates over $\Delta$ time units. The differential equation $d\boldsymbol{\mu}(t)=\boldsymbol{\nu}(t)dt$ represents the instantaneous velocity as $\Delta$ approaches zero under the assumption that $\boldsymbol\mu(t)$ is a continuous movement path [@johnson2008continuous]. Although the bivariate velocity process $\boldsymbol\nu(t) = [\nu_x(t), \nu_y(t)]'$ could be considered as cross-correlated between the axes, these correlated velocities produce strange directed travel and thus velocity in each coordinate is considered to be an independent process [@johnson2008continuous]. We decided instead, to focus on a velocity-based movement process using velocity scalars and turning angles. Under this framework, we define a scalar for instantaneous directionless velocity as $$\nu(t+\Delta) = \sqrt{(\mu_x(t+\Delta)-\mu_x(t))^2 + (\mu_y(t+\Delta) - \mu_y(t))^2}$$ and assume it to be a random variable with probability density $N \sim f(t|\boldsymbol\sigma)$ where $\boldsymbol\sigma$ represents the vector of parameters and determines the overall variability in velocity according to a specific probability density function $f(\boldsymbol\sigma)$. We assumed velocities to be independent of turning angles and chose to describe each with a separate probability density function, which is common practice when using movement lengths [@kareiva1983analyzing]. We calculated the turning angle $\theta(t+\Delta)$ describing the turning direction of the animal at time $t+\Delta$ and assumed the distribution of angles along a continuous movement path to be distributed with probability density $g(\theta)$.  For the specific case of a correlated random walk, the distribution of $g(\theta)$ can be used as a measure of correlation in the direction of movement, where if no correlation is present, the distribution is uniform [@kareiva1983analyzing].  As in other papers, we adopt the convention of using upper case letters for random variables and lower case letters for possible numerical values of these random variables.

Using this velocity process and turning angles, the continuous-time location process $\boldsymbol\mu(t)$ can be obtained by integrating over the velocity process [@johnson2008continuous]:
$$\boldsymbol\mu(t) = \boldsymbol\mu(0) + \int_0^t\boldsymbol\nu(u)du$$
where 
$$\int_0^t\boldsymbol\nu(u)du =\int_0^t [\nu_x(u), \nu_y(u)]' du = \int_0^t [\sin(\theta_u), \cos(\theta_u)]' \cdot \nu(u)du$$
This integral describes the location at time $t$ as the sum of steps determined by velocities in each direction plus a starting location at time $t=0$.

## Model formulations

## Model evaluation using information criteria

## Example: Many-banded aracari movement
